Community-conditioned language models for linguistic variation

### 2021-10-16 paper/paper1.tex
final edits for aclrr. add ling. sim/soc sim examples to fig 2; add some related work; ethics statement; small changes to intrudction, conclusion

% Remove the "review" option to generate the final version.
\usepackage[review]{emnlp2021}

@@

@@

\section{Introduction}
Linguistic communication requires that speakers share
certain linguistic conventions, such as syntactic
structure, word meanings, and patterns of interaction.
Speakers assume that these conventions are \emph{common ground} among
their interlocutors, based on joint membership in a community
\cite{Stalnaker2002, Clark1996}.  Such a community may be of any size,
from the very small, like members of a particular friend group, to the
very large, like speakers of English.  Because it is the source of
linguistic convention, the \emph{speech community} \citep{Gumperz1972}
is an important concept in the study of linguistic variation.

@@

\paragraph{Perplexity}

@@

\paragraph{CCLM Information Gain}

@@

\newcommand{\modelboxplot}[3]{
  \addplot+[
    boxplot={draw position=#3, box extend=0.3}, 
    draw=#2, mark=*, mark options={fill=#2, scale=0.5}, solid, fill=#2!10,
    area legend] 
    table [y=#1] {floats/comm.csv};
}
\begin{figure}[t]
\begin{tikzpicture}
  \begin{axis}[name=ig,
    height=7cm,width=8cm,
    boxplot/draw direction=x, axis x line*=bottom, axis y line*=left,y axis line style={opacity=0},
    xmax=1.6,
    ylabel=$l_c$, xlabel=$\IG_{M_j}$,
    legend style={font=\tiny,at={(1,1)},anchor=north east}, legend columns=2
    ]
    \addplot[thin, red, samples=2] coordinates {(1,-0.5)(1,3.5)};
    \foreach \lc in {0, ..., 3}{
      \modelboxplot{lstm-3-\lc-ig}{violet}{\lc+0.2}
      \modelboxplot{transformer-3-\lc-ig}{teal}{\lc-0.2}
    \legend{,LSTM, Transformer}
    }
  \end{axis}
  \begin{axis}[name=ppl,
    anchor=north west,at={($(ig.south west)-(0,1.1cm)$)},
    width=8cm, height=2.5cm, 
    boxplot/draw direction=x, axis x line*=bottom, hide y axis,
    xlabel=$\Ppl_{M_j}$,
    ]
    \modelboxplot{lstm-3-ppl}{violet}{0.2}
    \modelboxplot{transformer-3-ppl}{teal}{-.2}
  \end{axis}
\end{tikzpicture}
\caption{
  Average model performance by community.
  The boxes indicate upper and lower quartiles, while the whiskers are placed at the upper and lower maximum, with communities more than $1.5\times IQR$ (inter-quartile range) above the upper quartile considered outliers (represented as dots). The three most extreme outliers are excluded from this view (see \cref{fig:ppl-info-gain}).
   %See also table 1 in the supplementary materials % not very useful for this actually since we sort by indisc and it's only one model anyway.
   %for a tabular version of these results with community names.
}
\label{fig:comm-stratified-box}
\end{figure}

@@

To this end, we compare the CCLM-learned community
embeddings\footnote{
  In this section, we only consider the embeddings from 
  the \emph{best} (highest informaiton gain) CCLM from each architecture family; 
  that is, the LSTM with $l_c=1$ and the Transformer with $l_c=0$, however we
  observed similar results for other values of $l_c$.}
with the community embedding created by \citet{Kumar2018},
which were generated using using a negative-sampling optimization
algorithm, with the message author-community co-occurrence matrix as ground
truth, using data from January 2014 to April 2017.
We refer the reader
to \citet{Kumar2018}
for details, but the important point is that no linguistic
information is used to create these embeddings: they only reflect the social
relationship between communities, via 
the user-community membership relation. In contrast, CCLM community embeddings depend in no
way on which user is the author of any given message: we only use the
contents of messages, not authorship data.
%

@@

\begin{figure*}[ht]
\begin{tikzpicture}
  \begin{groupplot}[group style={
      group size=2 by 1,
      horizontal sep=1cm,
    },
    height=5cm,width=8cm,
    axis x line*=bottom, axis y line*=left,x axis line style={opacity=0},
    ymin=-0.35, ymax=1.00,
    xtick=data,
    xticklabel style={align=left, anchor=east, rotate=60},
    minor tick length=1ex,
    ]
    \nextgroupplot[ybar,
      bar width=0.25cm,
      xticklabels from table={floats/low_soc_high_ling.csv}{comms-label},
      ylabel={cosine similarity},
    ]
    \addplot[draw=violet,fill=violet!10] table [y=lstm-sim] {floats/low_soc_high_ling.csv};
    \addplot[draw=teal,fill=teal!10] table [y=transformer-sim] {floats/low_soc_high_ling.csv};
    \addplot[draw=red,fill=red!10] table [y=social-sim] {floats/low_soc_high_ling.csv};
    \nextgroupplot[ybar,
      bar width=0.25cm,
      xticklabels from table={floats/low_ling_high_soc.csv}{comms-label},
      legend style={font=\tiny,at={(1,1)},anchor=south east}, legend columns=3
    ]
    \addplot[draw=violet,fill=violet!10] table [y=lstm-sim] {floats/low_ling_high_soc.csv};
    \addplot[draw=teal,fill=teal!10] table [y=transformer-sim] {floats/low_ling_high_soc.csv};
    \addplot[draw=red,fill=red!10] table [y=social-sim] {floats/low_ling_high_soc.csv};
    \legend{LSTM, Transformer, Social}
  \end{groupplot}
\end{tikzpicture}
\caption{
  Communities with high linguistic similarity but low social similarity (left) and \textit{vice versa} (right). 
  Linguistic similarity is measured with the embeddings from the best LSTM ($l_c=1$) and Transformer ($l_c=0$)
  models and social similarity is measured with the embedding from \citet{Kumar2018}.
}
\label{fig:embedding-differences-examples}
\end{figure*}

@@

\todo{insert some examples with very high correlation}

The examples shown in \cref{fig:embedding-differences-examples} were
selected by taking the \num{5} most linguistically similar (in the
LSTM embedding) pairs of communities that had below median social
similarity (and \textit{vice versa}).\footnote{ Median similarity
  among pairs of communities was \num{0.177} for the social embedding
  and \num{0.010} and \num{0.012} for the LSTM and transformer
  linguistic embeddings, respectively.}

As one would expect, it seems that the communities with similar
linguistic embeddings but dissimilar social embeddings
(\cref{fig:embedding-differences-examples}, left) share a similar
topic, mode of interaction, or language variety, but for various
reasons, may attract a quite different set of members.  For example,
\subreddit{hiphopheads} and \subreddit{Monstercat} are both topically
related to music, but the music genres are different, and the later
has a more geographically local focus (Monstercat is an independent
electronic music label based in Vancouver).  While both
\subreddit{MLPLounge} and \subreddit{CasualConversation} are oriented
towards casual general-topic conversation, the former is geared
towards members of a niche internet
sub-culture. % adult fans of the television show My Little Pony. There are other subreddits (mentioned in r/MLPLounge's "about") that are for on-topic discussion of MLP
Both \subreddit{rocketbeans} and \subreddit{de} are primarily
German-language subreddits, but the former is comprised of fans of a
computer gaming-related YouTube channel, while the later is more
general-interest.

@@

Here, the orthogonal matrix \(Ω\) gives a map from linguistic embeddings to social embeddings.
The problem of computing \(d(L,S)\) is known as the orthogonal Procrustes problem~\citep{Gower2004}.\footnote{
This approach has also been used to compare word embeddings across representations \citep[e.g.,][]{Hamilton2016}.}
The solution is
\[d(L,S) = n - \mathsf {Tr} (Σ)\] where the matrix \(Σ\) is obtained by the singular value
decomposition (SVD) \(U^TΣV = LS^T\). Additionally the vectors of \(U\) and \(V\) give
the directions of correlation respectively of \(L\) and \(S\). That is, each
singular value \(\sigma_i\) (the elements of the diagonal matrix \(Σ\)), give a measure of how much
correlation there is between the directions \(U_i\) and \(V_i\).

@@

@@

Our main result is that the the representation of communities built by
CCLM has a positive correlation with user co-occurrence patterns.
Even though kind of correlation is a core hypothesis of
socio-linguistics,\todo{what was this effect called?}  we believe that
this study is the first to test it on topic-organised
communities. Furthermore, it appears that our method (correlating
linguistic embeddings and social embeddings) is novel. Indeed, even
though the procrustes method has been used to correlate two sets of
linguistic emdeddings \emph{for the same model}, we find no evidence
of the method being applied to embeddings for widely different models,
as we have done.

@@

@@

>>>>

% Remove the "review" option to generate the final version.
\usepackage[review]{acl}

@@

% If the title and author information does not fit in the area allocated, uncomment the following
%
%\setlength\titlebox{<dim>}
%
% and set <dim> to something 5cm or larger.

@@

\section{Introduction}

Linguistic communication requires that speakers share
certain linguistic conventions, such as syntactic
structure, word meanings, and patterns of interaction.
Speakers assume that these conventions are \emph{common ground} among
their interlocutors, based on joint membership in a community
\cite{Stalnaker2002, Clark1996}.  Such a \emph{speech community} \citep{Gumperz1972} 
may be of any size, from the very small, like members of a particular friend group, 
to the very large, like speakers of English; however as \citet{Eckert1992}
point out, it is \emph{communities of practice}---defined by mutual social engagement
in some common activity---that are the primary locus of linguistic variation.

@@

\subsection{Perplexity}\label{sec:ppl}

@@

\paragraph{CCLM Information Gain}\label{sec:ig}

@@

\newcommand{\modelboxplot}[3]{
  \addplot+[
    boxplot={draw position=#3, box extend=0.3}, 
    draw=#2, mark=*, mark options={fill=#2, scale=0.5}, solid, fill=#2!10,
    area legend] 
    table [y=#1] {floats/comm.csv};
}
\begin{figure}[t]
\begin{tikzpicture}
  \begin{axis}[name=ig,
    height=7cm,width=8cm,
    boxplot/draw direction=x, axis x line*=bottom, axis y line*=left,y axis line style={opacity=0},
    xmax=1.6,
    ylabel=$l_c$, xlabel=$\IG_{M_j}$,
    legend style={font=\tiny,at={(1,1)},anchor=north east}, legend columns=2
    ]
    \addplot[thin, red, samples=2] coordinates {(1,-0.5)(1,3.5)};
    \foreach \lc in {0, ..., 3}{
      \modelboxplot{lstm-3-\lc-ig}{violet}{\lc+0.2}
      \modelboxplot{transformer-3-\lc-ig}{teal}{\lc-0.2}
    \legend{,LSTM, Transformer}
    }
  \end{axis}
  \begin{axis}[name=ppl,
    anchor=north west,at={($(ig.south west)-(0,1.1cm)$)},
    width=8cm, height=2.5cm, 
    boxplot/draw direction=x, axis x line*=bottom, hide y axis,
    xlabel=$\Ppl_{M_j}$,
    ]
    \modelboxplot{lstm-3-ppl}{violet}{0.2}
    \modelboxplot{transformer-3-ppl}{teal}{-.2}
  \end{axis}
\end{tikzpicture}
\caption{
  Average model performance by community.
The boxes indicate upper and lower quartiles, while the whiskers are placed at the upper and lower maximum, with communities more than $1.5\times IQR$ (inter-quartile range) above the upper quartile considered outliers (represented as dots). The three most extreme outliers are excluded from this view.}.% \cref{fig:ppl-info-gain}).
   %See also table 1 in the supplementary materials % not very useful for this actually since we sort by indisc and it's only one model anyway.
   %for a tabular version of these results with community names.
\label{fig:comm-stratified-box}
\end{figure}

@@

To this end, we compare the CCLM-learned community
embeddings\footnote{
  In this section, we only consider the embeddings from 
  the \emph{best} (highest information gain) CCLM from each architecture family; 
  that is, the LSTM with $l_c=1$ and the Transformer with $l_c=0$, however we
  observed similar results for other values of $l_c$.}
with the community embedding created by \citet{Kumar2018},
which were generated using using a negative-sampling optimization
algorithm, with the message author-community co-occurrence matrix as ground
truth, using data from January 2014 to April 2017.
We refer the reader
to \citet{Kumar2018}
for details, but the important point is that no linguistic
information is used to create these embeddings: they only reflect the social
relationship between communities, via 
the user-community membership relation. In contrast, CCLM community embeddings depend in no
way on which user is the author of any given message: we only use the
contents of messages, not authorship data.
%

@@

\newcommand{\pairwisecossimbarplot}[1]{
    \nextgroupplot[ybar=0pt,
      bar width=0.25cm,
      xticklabels from table={floats/#1}{comms-label},
      enlargelimits=0.15,
      yticklabels={,,}
    ]
    \addplot[draw=violet,fill=violet!10] table [y=lstm-sim] {floats/#1};
    \addplot[draw=teal,fill=teal!10] table [y=transformer-sim] {floats/#1};
    \addplot[draw=red,fill=red!10] table [y=social-sim] {floats/#1};
}

\begin{figure*}[ht]
\begin{tikzpicture}
  \begin{groupplot}[group style={
      group size=3 by 1,
      horizontal sep=0.5cm,
    },
    height=5cm,width=0.39\textwidth,
    axis x line*=bottom, axis y line*=left,x axis line style={opacity=0},
    ymin=-0.35, ymax=1.00,
    xtick=data,
    xticklabel style={align=right, anchor=east, rotate=60},
    minor tick length=1ex,
    legend style={font=\tiny,at={(1,1)},anchor=north east}, legend columns=3
    ]
    \nextgroupplot[ybar=0pt,
      bar width=0.25cm,
      xticklabels from table={floats/high_ling_high_soc.csv}{comms-label},
      enlargelimits=0.15,
    ]
    \addplot[draw=violet,fill=violet!10] table [y=lstm-sim] {floats/high_ling_high_soc.csv};
    \addplot[draw=teal,fill=teal!10] table [y=transformer-sim] {floats/high_ling_high_soc.csv};
    \addplot[draw=red,fill=red!10] table [y=social-sim] {floats/high_ling_high_soc.csv};
    \pairwisecossimbarplot{high_ling_low_soc.csv}{}{}
    \pairwisecossimbarplot{low_ling_high_soc.csv}{}{}
    \legend{LSTM, Transformer, Social}
  \end{groupplot}
\end{tikzpicture}
\caption{
  Cosine similarity between pairs of communities, computed for vectors from the
  best CCLM embeddings (LSTM: $l_c=1$, Transformer: $l_c=0$)
  and the social embedding from \citet{Kumar2018}.
  Communities with high linguistic and social similarity \textbf{(left)},
  high linguistic but low social similarity \textbf{(center)}, and
  low linguistic but high social similarity \textbf{(right)}.
  See text for details on the selection criteria.}
  \label{fig:embedding-differences-examples}
\end{figure*}

@@

We consider communities from three different selection criteria: 
Those with high social \emph{and} linguistic similarity 
(where the sum of the two is highest),
those with high linguistic and low social similarity
(where social similarity is below the median and linguistic similarity is highest), and
those with low linguistic and high social similarity
(where linguistic similarity is below these median and linguistic similarity is highest).\footnote{We use the LSTM ($l_c=1$) community vectors for these purposes, 
  but results attain with the best transformer model.}\footnote{Median similarity
  among pairs of communities was \num{0.177} for the social embedding
  and \num{0.010} and \num{0.012} for the LSTM and transformer
  linguistic embeddings, respectively.}
We do not consider pairs of communities that are different in both ways,
since these don't offer much in the way of understanding the respective embeddings.

Unsuprisingly, the first category (\cref{fig:embedding-differences-examples}, left)yields communities that are qualitatively very similar. 
The \subreddit{SSBPM} and \subreddit{darksouls} communities are focused
around discussion of a particular videogame, and are paired
with communities that discuss a variation of the same game.
The \subreddit{amiugly} and \subreddit{Rateme} communities are both 
forums where the posts are selfies and the comments are mostly 
comments on the person's appearance.
The two communities paired with \subreddit{reddevils} are 
likewise comprised of fans of a particular English football club.
That the three football club communities are so closely related points
to an important point about the social embedding: Because of the way
it was trained, it does not have to be the case that communities with
a similar representation have a high overlap in their membership,
but rather that the \emph{kinds of members} of the first community
are also the \emph{kinds of members} who would belong to the second, 
based on their membership in other communities.

Communities with similar linguistic embeddings but dissimilar social embeddings
(\cref{fig:embedding-differences-examples}, left) tend to share a similar
topic, mode of interaction, or language variety, but 
in all cases we looked at, there is some reason to expect that
they might nevertheless attract different members.
For example,
\subreddit{hiphopheads} and \subreddit{Monstercat} are both topically
related to music, but the music genres are different, and the later
has a more geographically local focus (Monstercat is an independent
electronic music label based in Vancouver). The interactions in both 
\subreddit{MLPLounge} and \subreddit{CasualConversation} 
could be described as casual conversation, the former is intended specifically for
members of a niche internet sub-culture. % adult fans of the children's television show My Little Pony. There are other subreddits (mentioned in r/MLPLounge's "about") that are for on-topic discussion of MLP
The \subreddit{exmormon} and \subreddit{Catholicism} communities discuss
the Mormon and Catholic churches, although their members have different 
relationships towards those organizations---the former is intended for
former members of the church, whereas the later is geared towards practicing
Catholics.
Finally, both \subreddit{rocketbeans} and \subreddit{de} are primarily
German-language subreddits, but the former is comprised of fans of a
computer gaming-related YouTube channel, while the later is more
general-interest.

@@

Here, the orthogonal matrix \(Ω\) gives a map from linguistic embeddings to social embeddings.
The problem of computing \(d(L,S)\) is known as the orthogonal Procrustes problem~\citep{Gower2004}.\footnote{
This approach has also been used to compare word embeddings across representations \citep[e.g.,][]{Hamilton2016a}.}
The solution is
\[d(L,S) = n - \mathsf {Tr} (Σ)\] where the matrix \(Σ\) is obtained by the singular value
decomposition (SVD) \(U^TΣV = LS^T\). Additionally the vectors of \(U\) and \(V\) give
the directions of correlation respectively of \(L\) and \(S\). That is, each
singular value \(\sigma_i\) (the elements of the diagonal matrix \(Σ\)), give a measure of how much
correlation there is between the directions \(U_i\) and \(V_i\).

@@

There are several recent studies that aim to measure \emph{linguistic distinctiveness}
at the level of speech community, which is an possible interpretation of 
of community-stratified information gain the CCLM over its unconditioned counterpart \cref{sec:ig}. Whereas the metrics in previous work are based on lexical frequency
\citep{OConnor2010,Zhang2018,Lucy2021}, CCLM information gain is capable of capturing 
distinctiveness at other levels of linguistic analysis. Further work is needed to
investigate exactly what kinds of variation are captured.

%- DelTredici 2017
%- Einstein 2010
%- Baman et al 2014/2014a
%- Martin 2017 (community2vec), Kumar 2018
%- Huang 2014
%- Lucy 2021

@@

Our main result is that the representation of communities built by
CCLM has a positive correlation with user co-occurrence patterns.
Even though such \emph{homophilic} correlation is a core hypothesis of 
sociolinguistics (see \citet{Kovacs2020}, for example), 
we believe that this study is the first to test it at 
the level of communities of practice using computatial methods.
Furthermore, it appears that our method (correlating
linguistic embeddings and social embeddings) is novel. Indeed, even
though the procrustes method has been used to correlate two sets of
linguistic embeddings \emph{for the same model}, we find no evidence
of the method being applied to embeddings for widely different models,
as we have done.

@@

\section{Ethical considerations}

Our work uses pulicly available data from Reddit,
collected from the API made available by \citet{Baumgartner2020}.
Additional considerations apply, however 
(see \citet{Gliniecka2021} for discussion). Reddit users are not,
in general, aware of the possiiblity that their data will be used 
for research purposes, and deleted posts can persist in archive formats.
We are not reseasing any data or trained models as part of this study,
since the data is publicly available and our models have low enough
computational requirements as to pose little barrier for erplication.

In the qualitative discussion offered in \cref{sec:storytelling},
our comparative characterization of the topic, mode of interaction, 
and language varieties
used in the pairs of communities were formed by reading comments from the data our 
language models were trained on. This included Googling words and phrases that 
were unfamiliar. 
Where we make claims about the how the community is ``premised'' or what kinds of 
members it is ``geared towards'' or ``intended for'', these are based on the text
of the sidebar on the community's Reddit page.
While we believe this methodology, aggregated over many pairs of communities, 
is appropriate for making a qualitative comparison
of the community features encoded by different representations, 
to make conclusions about \emph{particular} communities based on such a cursory and
subjective analysis would be dubious and potentially harmful.

@@

\include{appendix}
